---
lang: en-US
bibliography: example_article.bib
csl: ../../share/pensoft-journals.csl
link-citations: true
colorlinks: true
params:
  volume: 15
  issue: 1
output:
  bookdown::html_document2:
    number_sections: false
    toc: false
    css: 
      - ../../share/style.css
  bookdown::epub_book:
    number_sections: false
    toc: false
    chapter_level: 1
    stylesheet: 
      - ../../share/style_epub.css
---

```{r, include=FALSE}
source("../../share/setup.R")
```

```{r, child="../../share/header_html.Rmd"}
```

# How to format an article for the *AKES Newsletter*

*by Matt Bowser^[USFWS Kenai National Wildlife Refuge, Soldotna, Alaska, <matt_bowser@fws.gov>]*

## Background

The first issue of the *Newsletter of the Alaska Entomological Society*, published in 2006, was written in Microsoft^®^ Word. To improve the appearance and uniformity of the *Newsletter*, from beginning in 2008 it was written in LaTeX^[<https://www.latex-project.org/>], borrowing most of its styling from the journal *R News*^[<https://cran.r-project.org/doc/Rnews/>]. In order to make formatting articles easier and in order to increase accessibility by adding an HTML output option, a version of the *Newsletter* was created in RStudio using R Markdown. Source code from 2020 to 2023 issues of the *AKES Newsletter* are available from a GitHub repository^[<https://github.com/Alaska-Entomological-Society/AKES_Newsletter>].

This article describes the way articles should be prepared for submission to the *AKES Newsletter*.

## Writing articles

To get started, download the example article release^[<https://github.com/Alaska-Entomological-Society/AKES_Newsletter/releases/download/example_v2023-11-29/example_article_2023-11-29.zip>] from GitHub, which includes 

- `year/issue/example_article.Rmd`, the R Markdown source of this example article;
- `year/issue/example_article.bib`, the bibliographic data for this article in  BibTeX format; 
- `year/issue/AKES_newsletter_year_issue.Rmd`, an example *AKES Newsletter* issue; and 
- `share/*`, shared files used for document setup and styling.

The `example_article.Rmd` R Markdown file can then be renamed as appropriate and edited in a text editor or in an integrated development environment like RStudio.  See @RStudio2014 and @Xieetal2023 for a guide to using R Markdown. See examples of previous articles from the 2022 and 2023 issues of the *Newsletter*. Source R Markdown files from all submitted articles are available from the GitHub repository. 

## Rendering

Individual articles can be opened in RStudio and rendered to HTML output by pressing the "Knit"  button. Rendering of individual articles should suffice for testing readiness for submission to the editors of the *Newsletter*.

To render the complete issue in Portable Document Format (PDF),

1. open the example *AKES Newsletter* issue R Markdown file in Rstudio,
2. edit the lines in the YAML header below `section-refs-bibliography:`, adding all of the bibliography files,
3. edit the `articles <- c("example_article.Rmd")` line so that all articles are included, and
4. press the "Knit" button in RStudio.

Rendering of the PDF version of the *Newsletter* requires many LaTeX packages as dependencies, so multiple packages may need to be installed before the *Newsletter* can be rendered.

## References

