---
lang: en-US
bibliography: example_article.bib
csl: ../../share/pensoft-journals.csl
colorlinks: true
params:
  volume: 15
  issue: 1
output:
  bookdown::html_document2:
    number_sections: false
    toc: false
    css: 
      - ../../share/style.css
  bookdown::epub_book:
    number_sections: false
    toc: false
    chapter_level: 1
    stylesheet: 
      - ../../share/style_epub.css
---

```{r, include=FALSE}
source("../../share/setup.R")
```

```{r, child="../../share/header_html.Rmd"}
```

# How to format an article for the *AKES Newsletter*

*by Matt Bowser^[USFWS Kenai National Wildlife Refuge, Soldotna, Alaska, <matt_bowser@fws.gov>]*

## Background

The first issue of the *Newsletter of the Alaska Entomological Society* (2006) was written in Microsoft^®^ Word. To improve the appearance and uniformity of the *Newsletter*, from 2008 to 2021 it was written in LaTeX^[<https://www.latex-project.org/>], borrowing most of its styling from the journal *R News*^[<https://cran.r-project.org/doc/Rnews/>]. In order to make formatting articles easier and in order to increase accessibility by adding an HTML output option, a version of the *Newsletter* was created in RStudio using R Markdown. Source code from 2020 to 2023 issues of the *AKES Newsletter* are available from a GitHub repository^[<https://github.com/Alaska-Entomological-Society/AKES_Newsletter>].

This article describes the way articles should be prepared and provides examples of typical figures, a table, etc.

## Writing articles

To get started, download the example repository from GitHub, which will include 

- `example_year/example_issue/example_article.Rmd`, the R Markdown source of this example article;
- `example_year/example_issue/example_article.bib`, the bibliographic data for this article in  BibTeX format; 
- `example_year/example_issue/AKES_newsletter_year_issue.Rmd`, an example *AKES Newsletter* issue; and 
- `share/*`, shared files used for document setup and styling.

Articles are written in R Markdown, a simple format that can be edited in a text editor or in an integrated development environment like RStudio. See @Xieetal2023 for a guide to using R Markdown.

An article that does not include any references only requires a single file; for an article with references, these must be provided as a text file in BibTeX format.



## References

