---
lang: en-US
bibliography: orchid_pollinators.bib
csl: ../../share/pensoft-journals.csl
link-citations: true
colorlinks: true
params:
  volume: 15
  issue: 1
output:
  bookdown::html_document2:
    number_sections: false
    toc: false
    css: 
      - ../../share/style.css
  bookdown::epub_book:
    number_sections: false
    toc: false
    chapter_level: 1
    stylesheet: 
      - ../../share/style_epub.css
---

```{r, include=FALSE}
source("../../share/setup.R")
```

```{r, child="../../share/header_html.Rmd"}
```

# New pollinators and insect visitors to orchids in Southeast Alaska and the Aleutian Islands, and components of the orchid pollinator network in Southeast Alaska

*by Marlin Bowles^[Juneau, Alaska, mbowles@mortonarb.org] and Bob Armstrong^[Juneau, Alaska, bob@discoverysoutheast.org]*

## Abstract

We provide new records for insects pollinating or foraging on orchids in Southeast Alaska as well as the Aleutian Islands.  Bumble bees (*Bombus* spp.) were observed foraging on the ordinarily self-pollinating *Platanthera aquilonis* and on the food-deceptive *Calypso bulbosa* var. *occidentalis*.  The latter species is thought to be pollinated by queen *Bombus* species, however we observed *B. melanopygus* workers due to a late spring.  We also found *B. melanopygus* pollinating *Goodyera oblongifolia*, and significant variation in pod production between two populations.  A Noctuidae moth, apparently *Mniotype tenera* was observed pollinating *Platanthera convallariifolia* in the Aleutian Islands.  Other Lepidoptera records include the Nocutuidae *Plusia* spp. on *Platanthera dilatata* and the Geometridae *Antepirrhoe fasciata* and *Rheumaptera subhastata* on *P. stricta*.  New Diptera records include mosquito (*Aedes*) species foraging and bearing pollinia on *P. obtusata* as well as the phorid fly *Megaselia* sp. and the dryomyza fly *Dryomyza anilis* foraging on *Neottia banksiana*. We suggest that this orchid may omit a fetid odor that attracts these carrion-feeding insects.  These records bring the number of insects pollinating or foraging on Southeast Alaska orchids to at least 20 species among 13 families in the Diptera, Hymenoptera, and Lepidoptera.  They also increase the number of orchids with verified pollinators in Southeast Alaska to 17 species. Most of these insects pollinate only 1 or 2 orchid species, indicating the importance of the entire insect community to the orchid population.  In contrast, *Bombus* species pollinate at least 5 orchids across the entire flowering season.  The spring-flowering *Calypso bulbosa* could be vulnerable to climate warming that causes this species to flower before emergence of queen bumblebees.

## Introduction

Given the threat of climate warming to disrupt plant-pollinator synchrony, especially spring flowering plants with specialized pollinators [@KudoIda2013], it is critical to understand these relationships within pollinator networks [@Schatzetal2020].  The Southeast Alaska archipelago and adjacent coastal area support at least 28 native orchid species and varieties [@BowlesArmstrong2019]; at least eight additional species occur in the remaining regions of Alaska.  The network of insects that pollinate these orchids is most well known for Southeast Alaska, where 15 pollinating insects have been identified for 14 orchid species, leaving half of the orchid flora with unknown pollinators  [@BowlesArmstrong2021].  This paper reports new information on pollinators and other insect visitors to orchids in Southeast Alaska, and one species from the Aleutian Islands. We also examine the relationships between known pollinator families and orchid genera, including several that may be vulnerable to climate warming.  

## Methods

With two exceptions, this work occurred within the City and Borough of Juneau, Alaska or on adjacent Tongass National Forest land. We used field observations and digital images for insect identification.  Confirmation was also accessed through online resources such as [BugGuide.net](https://bugguide.net), or specimen identification by an authority.  We recorded video and time-lapse images to determine presence of insect visitors to several species.  We also used bait to attract potential pollinators to one species, and an enclosed terrarium to observe potential pollinators of a second species.  Pollinators and visitors to eight orchid species are reported (Table 1).  All photos are by the authors except as indicated.

## Results

### *Calypso bulbosa* var. *occidentalis* (Linnaeus) Oakes (western fairy slipper)

This western variety reaches its northern range limit on forested islands in Southeast Alaska.  It is the first native orchid to flower and uses food deception to attract naive queen bumblebees (*Bombus* spp.), which soon switch to spring flowers that provide food resources [@Ackerman1981].  The lip of the single flower is a long-lasting showy slipper, and unpollinated plants may remain in flower for a month.  Unlike lady’s slipper orchids (*Cypripedium*), which retain pollinia and release pollen grains, entire pollinia are removed from *Calypso* by attachment to the scutellum, a smooth segment on the dorsal surface of the bee [@Ackerman1981].  The width of the gap between the orchid column (which holds pollinia) and the lip opening may range from 5 to 10 mm.  This variability, in combination with variability in bee species sizes and switching to flowers with food rewards, results in low seed pod production [@Ackerman1981].  *Salix* and *Vaccinium* species provided an alternative food source in Finland, where 7% of queen *Bombus* species carried *Calypso* pollinia while foraging on *Salix* [@Tuomietal2015], which is visited by *Bombus* species in our study area (Figure 1).

In 2023, a late spring in our study area delayed flowering of *C. bulbosa*, allowing overlap with emergence of *Bombus melanopygus* Nylander workers.  They were determined to be workers based on their small size and late emergence.  *Calypso bulbosa* plants translocated to the Jensen Arboretum, Juneau, were visited by these bumblebees (Figure 2), one of which entered a *C. bulbosa* flower (Figure 3).  Though we did not observe pollinia transfer, this is the first report of *Bombus* workers visiting *C. bulbosa* var. *occidentalis*, and demonstrates that a weather-caused shift in flowering may result in asynchrony with foraging queen bumblebees.  This orchid could be vulnerable to climate warming that would disrupt this synchrony, thereby reducing pollination and seed production.

### *Goodyera oblongifolia* Rafinesque (large rattlesnake plantain orchid)

This western orchid reaches its northern range limit in near-coastal forests of Southeast Alaska.  Extensive populations of plants averaging > 100 rosettes of vegetative and flowering plants per 10 m^2^ area occur where disturbance has reduced shrub cover.  Flowering occurs in late summer and extends into early fall.  The inflorescence comprises a spike of small white flowers that contain nectar and are reportedly pollinated by *Bombus* species that cause low (< 50%) seed pod production [@Ackerman1975].  Our previous monitoring of this species revealed nocturnal visitation by Noctuidae moths, which are apparently nectar thieves that do not contribute to reproduction [@BowlesArmstrong2021].  However, we were unable to document pollination by *Bombus* species.

In 2022, we documented flower visitation by bumblebees resembling *Bombus melanopygus* with time-lapse images and with direct observation.  Individuals bearing pollinia approached (Figure 4) and briefly visited (Figure 5) multiple flowers on an inflorescence before moving to neighboring plants.  They also appeared to forage especially on higher density patches, quickly moving through orchid populations.  We compared percent and total seed capsule production per inflorescence between two sites with different inflorescence densities and sizes, Rainforest and Tee Harbor (Figure 6).  Data are from *n* = 26 random 10 m^2^ plots, and from *n* = 109 random inflorescence samples.  Tee Harbor had 38% greater average percent seed capsule production, greater flowering plant density, and smaller inflorescences than did the Rainforest population.  Average total capsule production was also greater, by 33%, in the Tee Harbor population (*p* = 0.013, Mann-Whitney *t*-Test). This supports findings that greater flowering plant density enhances bee visitation and pollination at the local scale [@Hegland2014].  However, other factors, such as bee population size and nest proximity, as well as orchid population size, could affect visitation rates.  Nocturnal foraging by Noctuidae moths might also cause bumble bees to reject inflorescences that have been subject to high levels of nectar thievery [@BowlesArmstrong2021].  More work is needed to test this hypothesis.

### *Neottia* (*Listera*) *banksiana* (Lindley) Reichenbach (Northwest twayblade)

This western species reaches its northern range limit in coastal Southeast Alaska forests.  Pollination is required for seed pod production, which ranged from 10 to 80% in our study areas.  Flowering occurs in late spring, with an inflorescence of up to 20 small green flowers.  Most members of the genus are reported to provide small amounts of nectar at the base of the lip, and to attach pollinia to insects as they feed on the nectar.  The smaller heart-leaved twayblade  *Neottia cordata* (Linnaeus) Richards flowers earlier in similar habitat and is pollinated by fungus gnats, which are attracted to a strong fetid odor [@AckermanMesler1979].  *Dryomyza* sp. flies visited *N. banksiana* in previous years, but none were observed accessing nectar nor carrying pollinia [@BowlesArmstrong2021].  *Dryomyza anilis* Fallen feeds on salmon carcasses [@Hockingetal2009] as well as bear scat.  This leads us to hypothesize that *N. banksiana* produces a fetid smell to attract pollinators; however, it does not have a strong odor detectable to humans.  

In 2023, following @vanderNietetal2011, we placed pieces of salmon carcass as bait at the base of flowering *N. banksiana* plants to enhance attraction of potential pollinators.  This procedure attracted large numbers of *Dryomyza anilis* flies as well as the Phorid fly *Megaselia* sp.  Phorid flies also feed on salmon carcasses [@Hockingetal2009] and pollinate other plants [@Ruppetal2021].  Both insects were observed feeding on the lips of *N. banksiana*, but neither carried pollinia.  The *Megaselia* appears to be too small to contact the orchid column (Figure 7), but *Dryomyza* flies appear to be large enough to remove pollinia (Figure 8).  Additional work is needed to assess whether this orchid produces a fetid odor detectable by insects, and whether *Dryomyza* flies function as pollinators.

### *Platanthera aquilonis* Sheviak (northern green rein orchid)

This rein orchid was formerly treated with the tetraploid *P. hyperborea*, but was segregated as *P. aquilonis* [@Sheviak2001] as it is diploid and has a more western distribution in northern North America including southeast and south central Alaska.  These species self-pollinate or have rain-assisted pollination, but appear to provide nectar in short spurs and may have outcrossing races [@Sheviak2001; @Kropf2015].  \textit{Platanthera aquilonis} is also considered one of the diploid progenitors (x \textit{P. dilatata}) of the polyploid \textit{P. huronensis}; however, pollen vectors that would facilitate this cross are not well-known [@Wallace2023].  Though classified as a nocturnal moth-pollinated species [@HapemanInoue1997], @Kropf2015 observed *Bombus jonellus* Kirby visiting *P. hyperborea* in Iceland, which suggests that *Bombus* species might facilitate outcrossing in *P. aquilonis* as well. 

In 2019 we observed a Bombus species foraging on Platanthera aquilonis (Figure 9) in a mixed population of this species with P. dilatata and P. huronensis in an uplift meadow in the Haines area of Southeast Alaska.  This observation complements the finding of Kropf (2015).  Moreover, as bumble bees also appear to pollinate P. dilatata (Van der Voort et al., 2022), our observation provides further evidence that Bombus species may be pollen vectors between P. aquilonis and P. dilatata, allowing spontaneous populations of their hybrid P. huronensis.  More work on potential for pollinia transfer in P. aquilonis is needed to verify this process.


## References